---
title: "House Price Prediction Analysis"
author: "Jai Patel"
date: "2024-12-10"
output:
  pdf_document:
    citation_package: biblatex
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: 3
bibliography: ../References
csl: apa-old-doi-prefix.csl 
link-citations: yes
---

```{r Libraries, include=FALSE}
    library(tidyverse)
    library(lubridate)
    library(knitr)
    library(formatR)
    library(jpeg)
    library(png)
    opts_chunk$set(tidy=TRUE, tidy.opts = list(width.cutoff=60))

```

# Abstract

This project looks at house sales in King County, USA, from May 2014 to May 2015 to understand what affects home prices. The data includes information about house square feet, condition, number of bedrooms, and more. The main finding is that bigger homes, especially those with more living space, usually cost more. The year a house was built doesn’t have much effect on its price. The size of the above-ground space is important for price. Lot and basement size differ from one another. As the number of bathrooms increase, the price also goes up. Homes with more bedrooms tend to have better condition ratings. Even though the dataset has some limitations, the analysis gives useful insights into what influences home prices in the area.

# Introduction

This project focuses on exploring housing trends using this dataset. Large datasets like this are important for understanding the real estate market, planning cities, and making financial decisions. As housing demand increases, it’s important to use data to better understand market patterns and trends.

Researchers have used similar datasets in the past to study pricing trends, identify key property features, and predict how the market would change. However, new uses like machine learning features have improved visualization methods and offer a new way to see the change. This project will analyze the dataset to find patterns that affect housing prices, helping people make smarter decisions.

By examining factors like square feet of the house and condition, this project aims to provide clear, data-driven insights for real estate and property management. The goal is to use models and visualizations to turn this data into practical, useful information.

# Data

The dataset contains 21,613 rows and 22 columns, containing many variables. The variables are `X`, `id`, `date`, `price`, `bedrooms`, `bathrooms`, `sqft_living`, `floors`, `waterfront`, `view`, `condition`, `grade`, `sqft_above`, `sqft_basement`, `yr_built`, `yr_renovated`, `zipcode`, `lat`, `long`, `sqft_living15`, and `sqft_lot15` (see @Shiv).
The dataset focuses on house sales in King County, USA, including the city of Seattle. It covers transactions from May 2014 to May 2015, offering a idea of the housing market during that period. However, the dataset has some limits. It only covers one year, so it can’t show long-term trends. It also focuses only on King County, so the results may not apply to other areas. Even with these challenges, the dataset is a good starting point for understanding what impacts home prices.

```{r}

data= read.csv("../Documents/kc_house_data_NaN.csv")
summary(data)
```

The above data is showing basic information of the categories that we can potentially see a relationship between.

# Analysis

## Price & Square Foot in Living Relationship

```{r}
library(tidyverse)
library(ggplot2)

ggplot(data, aes(x = sqft_living, y = price)) +
    geom_point(color="green") +
  labs(
    title = "Relationship Between Price and Living Area",
       x = "Living Area",
       y = "Price") 

correlation = cor(data$sqft_living, data$price, use = "complete.obs")
paste0("Correlation is: ", correlation)



```

### Result

The scatter plot (see @R2021) shows that bigger houses with more square feet of living space usually cost more. There’s a upward trend, but there are a few exceptions where smaller houses have high prices, probably because of things like location or special features like have a view with water.

The correlation (see @RStudio2020) between the size of the living area and price is 0.7020350546118, which means there’s a strong positive relationship. This shows that as the size of the living area increases, the price also tends to go up.

This means bigger homes are generally more expensive, which makes sense in the housing market. This information is useful for understanding how size affects home prices.

## Price & Year Built Relationship

```{r}

model = lm(price ~ yr_built, data = data)

summary(model)
ggplot(data, aes(x = yr_built, y = price)) +
  geom_point(color = "red") +
  geom_smooth(method = "lm") +
  labs(title = "Price vs Year Built",
       x = "Year Built",
       y = "Price") 
```

### Result

The model shows a very weak link between the year a house was built and its price. On average, each additional year increases the price by \$675.10. However, this number is  small compared to other factors like the house size.

The R-squared value is 0.002917, meaning only about 0.2917% of the variation in house prices can be explained by the year the house was built. This suggests that other factors like size and condition have a much greater impact on the price. 

The scatter plot shows that newer houses are a bit more expensive, but the trend is very weak. The line in the plot is almost flat, and there’s a lot of scatter around it, saying that the year built doesn’t have a strong influence on price by itself.

## Square Foot Above and Price Relationship

```{r}

model = glm(price ~ sqft_above, data = data, family = gaussian())

summary(model)

plot(data$sqft_above, data$price, 
     xlab = "Square Footage Above Ground", ylab = "House Price", 
     main = "Price vs Square Foot Above Relationship")
abline(model, col = "blue")

```

### Result

In this analysis, the relationship between the square footage of above-ground space  and house price. The results show a strong positive correlation, where each additional square foot of above-ground space is associated with an increase in house price by \$268.50. The model's statistical significance, shown by the very low p-value (\< 2e-16), shows that square footage is a meaningful predictor of house prices.

## Floors + Bathrooms and Price Relationship

```{r}
library(ggplot2)

ggplot(data, aes(x = bathrooms, y = price, color = factor(floors))) +
  geom_point() +
  labs(
    title = "Price vs Bathrooms with Floors as a Factor",
       x = "Number of Bathrooms",
       y = "Price",
       color = "Number of Floors")

```

### Result

The analysis shows a positive relationship between the number of bathrooms and house price—meaning that as the number of bathrooms increases, the price tends to go up as well. The number of floors appears to be spread out across the graph, with no clear trend, but it seems to increase in the upward direction.

## Square Foot Basement and Square Foot Lot Relationship

```{r}

model = lm(sqft_basement ~ sqft_lot, data = data)
summary(model)

ggplot(model, aes(x = sqft_lot, y = sqft_basement)) +
  geom_point() +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Relationship Between Lot Size and Basement Size",
    x = "Lot Size",
    y = "Basement Size")

```

### Result

The analysis looks at the connection between lot size and basement size. It shows a very small positive trend, where basement size increases by about 0.00016 sqft for each additional square foot of lot size. While the p-value of 0.0246 means the relationship is statistically significant, the R-squared value is only 0.023%, so the model explains very little of the changes in basement size. The graph is mostly showing that the smaller the lot size, the bigger the basement size is. 

## Condition and Bedroom Relationship

```{r}

mod = aggregate(bedrooms ~ condition, data, mean)

barplot = ggplot(mod, aes(x = factor(condition), y = bedrooms)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  labs(
    title = "Number of Bedrooms by the Condition of the House",
    x = "Condition",
    y = "Number of Bedrooms")

barplot


```

### Result

The barplot reveals that as the number of bedrooms increases, the average condition rating (on a scale of 1 to 5) also improves. This could indicate that houses with more bedrooms are often better constructed or maintained. However, it’s important to interpret this trend cautiously, as other factors, such as location or view might also influence the condition ratings.

# Conclusion

In conclusion, this project helps us understand housing prices by looking at a dataset of house sales in King County, USA, from May 2014 to May 2015. The analysis shows that larger homes, with more square footage, generally have higher prices, with a clear connection between size and cost. On the other hand, the year a house was built has very little effect on its price. The size of the above-ground space, however, does impact the price, with each additional square foot increasing the price. The relationship between lot size and basement size is that the smaller the lot size, the bigger the basement size. Also, houses with more bedrooms tend to have better condition ratings. While the dataset is limited in terms of time and location, the findings provide useful insights into the factors that influence housing prices.
